---
title: "R Notebook"
output: 
  html_document: 
    self_contained: no
---

Fundamentally, Elo ratings are quite simple. All teams (or competitors for individual competition such as tennis), start with a baseline value e.g. 1500. The baseline is what an average team could expect their Elo value to be. 
The beautiful thing about Elo scores, is that they translate to expected outcomes. 
The starting Elo formula for calculating the expected outcome of a game is

$$
\text{Expected Score} = \frac{1}{1+10^{\frac{(Team_{B} - Team_{A})}{\text{Scaling Factor}}}}
$$

This looks very similar to the logistic function, e.g.

$$
f(x) = \frac{e^{\alpha + \beta X}}{1 + e^{\alpha + \beta X}}
$$

Which becomes more apparent as we look how Elo works.

- Expected Score is the win probability for $Team_{A}$
- $Team_{A}$ and $Team_{B}$ are the two teams playing
- Scaling Factor is an arbitary value that scales the values. Traditionally 400 is used, which means a difference in 400 Elo would give the favoured team a $90\%$ of winning.

For example, say $Team_{A}$ has an Elo rating of $1600$ and $Team_{B}$ has a rating of $1500$, then we simple input the values into the formula, giving $Team_{A}$ a $64\%$ chance of winning.

$$
\text{Expected Score/Win Probability} = \frac{1}{1+10^{\frac{(1500 - 1600)}{400}}} = \frac{1}{1 + 10^{\frac{-100}{400}}} = 0.64
$$

This can be done in R or Python with 

```{r}
get_expected_score <- function(team_a, team_b){
  exp <- (team_b - team_a) / 400
    return(1 / (1 + 10^exp))
}

#get_expected_score(team_a = 1500, team_b = 1500)
```

```{python}
def get_expected_score(team_a, team_b):
    exp = (team_b - team_a) / 400
    return(1 / (1 + 10**exp))
```


After each round or game, Elo values are adjusted depending if a team wins or loses. The amount values change after a single game is determined with $K$. 

First each teams win probability is calculated like above
Then the difference between their win probability (or expected score) and actual scores are used to make an adjustment
To do this, games are scored as $0$ for a loss, $0.5$ for a tie, and $1$ for a win. 
Once we have the difference between the expected value and game result, we multiple that value by our $K$ value. 
A larger $K$ creates more variance as the values get updated, whereas a lower $K$ value takes longer to adjust to new information. 538 in their Elo model use a $K$ of $25$.


$$
\underbrace{Elo_{Team_{A}}}_{\text{new Elo}} = \underbrace{Elo_{Team_{A}}}_{\text{old Elo}} + \text{ K Factor } \Big (\text{ Actual Score }_{Team_{A}} - \text{ Expected Score }_{Team_{A}}\Big )
$$

For example, imagine $Team_{A}$ and $Team_{B}$ play each other in round 1 with an Elo of 1500. Their expected score will both be 0.5 since they have the same Elo. $Team_{A}$ beats $Team_{B}$, which leads to $Team_{A}$ being assigned a $1$ for the victory and $Team_{B}$ a $0$. As such we can do

For $Team_{A}$

$$
\underbrace{Elo_{Team_{A}}}_{\text{new Elo}} = \underbrace{Elo_{Team_{A}}}_{\text{old Elo}} + 25 \times (1 - 0.5) = 25
$$

$$
Elo_{Team_{A}} = 1500+25 = 1525
$$

and $Team_{B}$

$$
\underbrace{Elo_{Team_{B}}}_{\text{new Elo}} = \underbrace{Elo_{Team_{B}}}_{\text{old Elo}} + 25 \times (0 - 0.5) = -12.5
$$

$$
Elo_{Team_{B}} = 1500-12.5 = 1487.5
$$


Calculating the new elo

```{r}
get_new_elos <- function(home_rating, away_rating, margin, k = 25) {
  
  home_score <- case_when(
    # 0 for a loss
    margin < 0 ~ 0,
    # 1 for a win
    margin > 0 ~ 1,
    # 0.5 if tie
    TRUE ~ 0.5
  )

  # get expected home score
  expected_home_score <- get_expected_score(home_rating, away_rating)
  
  # multiply difference of actual and expected score by k value and adjust home rating
  new_home_score <- home_rating + k * (home_score - expected_home_score) %>% round(2)

  # repeat these steps for the away team away score is inverse of home score
  away_score <- 1 - home_score
  expected_away_score <- get_expected_score(away_rating, home_rating)
  new_away_score <- away_rating + k * (away_score - expected_away_score) %>% round(2)

  # return a tuple
  return(new_home_score, new_away_score)
}
```

```{r}
games = c()

for year in range(2000, 2022):
    response = games_api.get_games(year=year)
    games = [*games, *response]

games = [dict(
            start_date=g.start_date,
            home_team=g.home_team,
            home_conference=g.home_conference,
            home_points=g.home_points,
            away_team=g.away_team,
            away_conference=g.away_conference,
            away_points=g.away_points
            ) for g in games if g.home_points is not None and g.away_points is not None]
games.sort(key=date_sort)
```



```{python}
def get_new_elos(home_rating, away_rating, margin):
    k = 25

    # score of 0.5 for a tie
    home_score = 0.5
    if margin > 0:
        # score of 1 for a win
        home_score = 1
    elif margin < 0:
        #score of 0 for a loss
        home_score = 0

    # get expected home score
    expected_home_score = get_expected_score(home_rating, away_rating)
    # multiply difference of actual and expected score by k value and adjust home rating
    new_home_score = home_rating + k * (home_score - expected_home_score)

    # repeat these steps for the away team
    # away score is inverse of home score
    away_score = 1 - home_score
    expected_away_score = get_expected_score(away_rating, home_rating)
    new_away_score = away_rating + k * (away_score - expected_away_score)

    # return a tuple
    return (round(new_home_score), round(new_away_score))
```



```{r}

```

```{r}

```

```{r}

```

```{r}
#https://blog.collegefootballdata.com/talking-tech-elo-ratings/
```

Just compare 10 games for the spread using 538 

```{r}
nfl_elo_538 <- read_csv("Downloads/nfl_new/nfl_elo_latest.csv") %>%
  filter(date == "2022-01-09") %>%
  select(home_team = team1, away_team = team2, home_elo = elo1_pre, away_elo = elo2_pre) %>%
  mutate(
    elo_diff = home_elo - away_elo,
    #round to nearest half
    predicted_spread = round((elo_diff/25) * 2)/2
  )



schedules <- nflreadr::load_schedules(seasons = 2021) %>%
  filter(gameday == "2022-01-09") %>%
  select(home_team, away_team, spread_line)


test <- nfl_elo_538 %>%
  left_join(schedules, by = c("home_team", "away_team"))
```

